Physics inspired priors for neutron star mergers improves equation of state constraints

### 2025-02-20 paper.tex
added model selection segment

	\begin{abstract}
		Gravitational-wave astronomy shows great promise in determining nuclear physics in a regime not accessible to terrestrial experiments. 
		We introduce physics-inspired priors derived from a combination of constraints from nuclear-physics experiments, perturbative quantum chromodynamics calculations, and astrophysical constraints.
		When these priors are used in gravitational-wave astrophysical inference, we show significant improvement on nuclear equation of state constraints. 
		Applying these to the first two gravitational-wave binary neutron star mergers GW170817 and GW190425, constraints on the radius of a $1.4\,M_\odot$ neutron star improves from $R=\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ to $\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ (90\% confidence intervals).
		We also show these priors can be used to perform model selection between binary neutron star and neutron star-black hole mergers, although the results for GW190425 are incolnclusive with a Bayes factor of $\red{}$ in favour of the binary neutron star merger hypothesis, representing only marginal evidence.
		We advocate for these physics-inspired priors to be used as standard in the literature, and provide open-source code for reproducibility and further adaptation of the method.
	\end{abstract}

@@

	\subsection{Binary neutron star mergers}
	We use the probability distribution of $10^7$ equation of state models informed by nuclear physics, perturbative quantum chromodynamics and  astrophysics from~\citet{altiparmak22} to generate a probability distribution for chirp mass -- tidal deformability, and use it as priors on those quantities as shown in Fig.~\ref{fig:priors}. We implement this in \textsc{Bilby} using the \verb!conditional_prior! function\footnote{\red{git link to our code!?}},
	interpolating the $200\times200$ grid to draw from a joint prior on $\mathcal{M}$ and $\tilde{\Lambda}$. 
	\begin{figure*}
		\centering
		\includegraphics[width=1.\linewidth]{Fig_1_Physics_priors.pdf}
		\caption{Coupled priors for the source-frame chirp mass $\mathcal{M}$ and dimensionless tidal deformability $\tilde{\Lambda}$. Top: prior for binary neutron star mergers. Bottom: prior for neutron star -- black hole mergers. The horizontal grey bands are the 90\% credible intervals for the measured chirp masses of GW170817~\citep{abbott17_170817observation}, GW190425~\citep{abbott20_190425}, GW200115~\citep{GW200115} and GW230529~\citep{GW230529}}. 
		\label{fig:priors}
	\end{figure*}

@@

	Figure \red{XX} shows the reconstructed mass---radius curves of GW170817 ... 
	\subsection{GW190425}
	Figure \ref{fig:GW190425_tides} shows the posterior distribution of the intrinsic tidal parameter of GW190425 when analyzed using both the binary neutron star prior (teal) and neutron star -- black hole prior (pink). We show the official posterior distribution  of the LVK obtained using uniform priors in blue. 
	Bayesian model selection between the binary neutron star and neutron star -- black hole merger gives a Bayes factor of \red{XX}, slightly in favor of a binary neutron star merger origin. 
	Figure \red{XX} shows the reconstructed mass radius curves of GW190425 analyzed using the new priors.  
	\begin{figure}
		\centering
		\includegraphics[width=1.\linewidth]{Fig_2_GW170817_lambda_tilde_posteriors_BNS.pdf}
		\caption{The posterior distributions of the tidal deformability $\tilde{\Lambda}$ obtained from Bayesian parameter estimation for GW170817. The posterior using a nuclear, microphysics, and astrophysics inspired prior is shown in teal, while the posterior using uniform priors from the LVK is shown in midnight blue. We show the (marginalized) prior with the light blue curve. Note, the posterior samples from the LVK show support up to $\tilde{\Lambda}$ = 2000. }
		\label{fig:GW170817_tides}
	\end{figure}

@@

	We generate two relations, one with uniform priors on our population of BH - NS mergers, and one directly informed by state of the art binary population synthesis. We use the mass ratio distribution of \cite{2021Broekgaarden+}  (\red{ what particular model are we using?}) for the generation of our population synthesis informed priors. Figure \ref{fig:mass ratio distrbution} shows the distribution in mass ratio used for the generation of the prior. 
	\begin{figure}
		\centering
		\includegraphics[width=\linewidth]{NSBH_mass_ratio_density.pdf}
		\caption{Probability density function of the mass ratio for detectable neutron star black hole mergers from \citep{2021Broekgaarden+}.}
		\label{fig:mass ratio distrbution}
	\end{figure}
	\section{Model Selection}\label{sec:Model Selection}

	\bibliography{references}{}
	\bibliographystyle{aasjournal}

>>>>

	\begin{abstract}
		Gravitational-wave astronomy shows great promise in determining nuclear physics in a regime not accessible to terrestrial experiments. 
		We introduce physics-inspired priors derived from a combination of constraints from nuclear-physics experiments, perturbative quantum chromodynamics calculations, and astrophysical constraints.
		When these priors are used in gravitational-wave astrophysical inference, we show significant improvement on nuclear equation of state constraints. 
		Applying these to the first two gravitational-wave binary neutron star mergers GW170817 and GW190425, constraints on the radius of a $1.4\,M_\odot$ neutron star improves from $R=\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ to $\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ (90\% confidence intervals).
		We also show these priors can be used to perform model selection between binary neutron star and neutron star-black hole mergers, although the results for GW190425 are incolnclusive with a Bayes factor of 1.33 in favour of the binary neutron star merger hypothesis, representing only marginal evidence.
		We advocate for these physics-inspired priors to be used as standard in the literature, and provide open-source code for reproducibility and further adaptation of the method.
	\end{abstract}

@@

	\subsection{Binary neutron star mergers}
	We use the probability distribution of $10^7$ equation of state models informed by nuclear physics, perturbative quantum chromodynamics and  astrophysics from~\citet{altiparmak22} to generate a probability distribution for chirp mass -- tidal deformability, and use it as priors on those quantities as shown in Fig.~\ref{fig:priors}. We implement this in \textsc{Bilby} using the \verb!conditional_prior! function\footnote{\red{git link to our code!?}},
	interpolating the $200\times200$ grid to draw from a joint prior on $\mathcal{M}$ and $\tilde{\Lambda}$. 
	\begin{figure}
		\centering
		\includegraphics[width=1.\linewidth]{Fig_1_Physics_priors.pdf}
		\caption{Coupled priors for the source-frame chirp mass $\mathcal{M}$ and dimensionless tidal deformability $\tilde{\Lambda}$. Top: prior for binary neutron star mergers. Bottom: prior for neutron star -- black hole mergers. The horizontal grey bands are the 90\% credible intervals for the measured chirp masses of GW170817~\citep{abbott17_170817observation}, GW190425~\citep{abbott20_190425}, GW200115~\citep{GW200115} and GW230529~\citep{GW230529}}. 
		\label{fig:priors}
	\end{figure}

@@

	Figure \red{XX} shows the reconstructed mass---radius curves of GW170817 ... 
	\subsection{GW190425}
	Figure \ref{fig:GW190425_tides} shows the posterior distribution of the intrinsic tidal parameter of GW190425 when analyzed using both the binary neutron star prior (teal) and neutron star -- black hole prior (pink). We show the official posterior distribution  of the LVK obtained using uniform priors in blue. 
	% Z_BNS = -576535.277281933
	% Z_NSBH = -576535.5652987232
	Bayesian model selection between the binary neutron star and neutron star -- black hole merger gives a Bayes factor of 1.33, marginally in favor of a binary neutron star merger origin. 
	Figure \red{XX} shows the reconstructed mass radius curves of GW190425 analyzed using the new priors.  
	\begin{figure}
		\centering
		\includegraphics[width=1.\linewidth]{Fig_2_GW170817_lambda_tilde_posteriors_BNS.pdf}
		\caption{The posterior distributions of the tidal deformability $\tilde{\Lambda}$ obtained from Bayesian parameter estimation for GW170817. The posterior using a nuclear, microphysics, and astrophysics inspired prior is shown in teal, while the posterior using uniform priors from the LVK is shown in midnight blue. We show the (marginalized) prior with the light blue curve. Note, the posterior samples from the LVK show support up to $\tilde{\Lambda}$ = 2000. }
		\label{fig:GW170817_tides}
	\end{figure}

@@

	We generate two relations, one with uniform priors on our population of BH - NS mergers, and one directly informed by state of the art binary population synthesis. We use the mass ratio distribution of \cite{2021Broekgaarden+}  (\red{ what particular model are we using?}) for the generation of our population synthesis informed priors. Figure \ref{fig:mass ratio distrbution} shows the distribution in mass ratio used for the generation of the prior. 
	\begin{figure}
		\centering
		\includegraphics[width=\linewidth]{NSBH_mass_ratio_density.pdf}
		\caption{Probability density function of the mass ratio for detectable neutron star black hole mergers from \citep{2021Broekgaarden+}.}
		\label{fig:mass ratio distrbution}
	\end{figure}
	\section{Model Selection}\label{sec:Model Selection}
	We perform parameter estimation of gravtitational-wave strain data using the \textsc{dynesty} nested sampler. Unlike standard MCMC samplers,
	\textsc{dynesty} calculates the evidence for each parameter estimation run.   
	To perform Bayesian model selection between binary neutron star and neutron star-- black hole mergers, we calculate the Bayes factor
	\begin{equation}
		\rm BF  = \frac{\mathcal{Z_{\rm BNS}}}{\mathcal{Z_{\rm NSBH}}},
	\end{equation}
	where $\mathcal{Z}_{\rm BNS}$ and $\mathcal{Z}_{\rm NSBH}$ are the Bayesian evidences obtained for the parameter estimation runs using the binary neutron star 
	and neutron star black hole priors respectively.
	%  In practice we compute the log Bayes factor
	% \begin{equation}
	% 	\log \ (\rm BF) = \log (Z_{\rm BNS} ) - \log (Z_{\rm NSBH}).
	% \end{equation} 
	Formally, for model selection we should be computing the odds ratio, which is defined as
	\begin{equation}
		\mathcal{O} \equiv \frac{Z_{\rm BNS}}{Z_{\rm NSBH}} \frac{\pi_{\rm BNS}}{\pi_{\rm NSBH}},
	\end{equation}
	where $\pi_{\rm BNS}$ and $\pi_{\rm NSBH}$ are the prior odds for each hypothesis; i.e our relative belief about which hypothesis is more likely. 
	In this work we set the prior odds to unity, so the odds ratio is the Bayes factor. 
	We could in principle construct an odds ratio where the prior odds was informed by the relative rates of binary neutron star and neutron -- star black hole mergers. 
	However since these rates are still highly uncertain, we choose the more agnostic approach of setting the prior odds to unity. 
	\bibliography{references}{}
	\bibliographystyle{aasjournal}
